---
title: "CPLN_Final_RB"
author: "Richard Barad"
date: "2024-04-24"
output: html_document
---

```{r setup, echo=FALSE}

library(sf)
library(tidycensus)
library(terra)
library(tidyverse)
library(tidyterra)
library(FNN)
library(gridExtra)
library(viridis)

library(caret)
library(yardstick)
library(pscl)
library(plotROC) 
library(ggrepel)
library(pROC)

palette2 <- c("#41b6c4","#253494")
palette4 <- c("#a1dab4","#41b6c4","#2c7fb8","#253494")
palette5 <- c("#ffffcc","#a1dab4","#41b6c4","#2c7fb8","#253494")
palette10 <- c("#f7fcf0","#e0f3db","#ccebc5","#a8ddb5","#7bccc4",
               "#4eb3d3","#2b8cbe","#0868ac","#084081","#f7fcf0")

```

```{r functions, echo=FALSE}

xyC <- function(aPolygonSF) {
  as.data.frame(
    cbind(x=st_coordinates(st_centroid(aPolygonSF))[,1],
          y=st_coordinates(st_centroid(aPolygonSF))[,2]))
} 

nn_function <- function(measureFrom,measureTo,k) {
  #convert the sf layers to matrices
  measureFrom_Matrix <-
    as.matrix(measureFrom)
  measureTo_Matrix <-
    as.matrix(measureTo)
  nn <-   
    get.knnx(measureTo, measureFrom, k)$nn.dist
    output <-
    as.data.frame(nn) %>%
    rownames_to_column(var = "thisPoint") %>%
    gather(points, point_distance, V1:ncol(.)) %>%
    arrange(as.numeric(thisPoint)) %>%
    group_by(thisPoint) %>%
    summarize(pointDistance = mean(point_distance)) %>%
    arrange(as.numeric(thisPoint)) %>% 
    dplyr::select(-thisPoint) %>%
    pull()
  
  return(output)  
}

quintileBreaks <- function(df,variable) {
    as.character(quantile(df[[variable]],
                          c(.01,.2,.4,.6,.8),na.rm=T))
}
```

# Get Data

## Get MSA for Atlanta

```{r atl_msa}

atl_msa <- st_read('https://arcgis.atlantaregional.com/arcgis/rest/services/OpenData/FeatureServer/67/query?outFields=*&where=1%3D1&f=geojson') %>%
  st_make_valid() %>%
  st_transform('EPSG:2240') %>%
  dplyr::filter(MSA == 'Y')

atl_msa_dissolve <- st_union(atl_msa)

```

## Get Land Cover Data

Import raster data and project it into the appropriate cordinate system. 

```{r atl_landcover}

lc2011 <- rast('DATA/nlcd_2011.tif')
lc2011 <- project(lc2011,'EPSG:2240')

lc2021 <- rast('DATA/nlcd_2021.tif')
lc2021 <- project(lc2021,'EPSG:2240')

```

# Create Fishnet for 2011

I create a 2500 x 2500 square foot grid.

```{r create_fishnet}

fishnet <- st_make_grid(atl_msa,2500) %>%
  st_sf() %>%
  st_join(.,atl_msa,predicate='intersects',left=FALSE) %>%
  mutate(ID=row_number()) %>%
  select(ID)
 
ggplot()+
  geom_sf(data=fishnet,color='grey50',fill='gray90',linewidth=0.1)+
  geom_sf(data=atl_msa,fill='transparent',color='black',linewidth=0.5)+
  theme_void()

```


# Reclassify Land Cover Rasters

Reclassify the land cover rasters into six categories: 'Water','Developed','Forest','Farm','Wetlands','OtherUndeveloped'

```{r reclassify_dev, fig.width=13, fig.height=7, warning = FALSE, message = FALSE}

land_cover_reclass_matrix = matrix(
                           c(11,0,
                             21,1,
                             22,1,
                             23,1,
                             24,1,
                             41,2,
                             42,2,
                             43,2,
                             81,3,
                             82,3,
                             90,4,
                             95,4,
                             52,5,
                             71,5,
                             31,5),
                             ncol=2,byrow = TRUE)

lc2011_class <- terra::classify(lc2011,land_cover_reclass_matrix)

levels(lc2011_class) <- c('Water','Developed','Forest','Farm','Wetlands','OtherUndeveloped')

lc2011_class_mask <- terra::mask(lc2011_class,atl_msa)

levels(lc2011_class_mask) <- c('Water','Developed','Forest','Farm','Wetlands','OtherUndeveloped')

lc2021_class <- terra::classify(lc2021,land_cover_reclass_matrix)

levels(lc2021_class) <- c('Water','Developed','Forest','Farm','Wetlands','OtherUndeveloped')

lc2021_class_mask <- terra::mask(lc2021_class,atl_msa)

levels(lc2021_class_mask) <- c('Water','Developed','Forest','Farm','Wetlands','OtherUndeveloped')

grid.arrange(ncol=2,

ggplot()+
  geom_spatraster(data=lc2011_class_mask,na.rm=TRUE)+
  geom_sf(data=atl_msa,fill='transparent',color='black',linewidth=0.5)+
  scale_fill_manual(values=c('lightblue','pink','lightgreen','lightyellow','blue','orange'),na.value = "transparent",name='Land Cover')+
  theme_void()+
  ggtitle('Land Cover in 2011'),

ggplot()+
  geom_spatraster(data=lc2021_class_mask,na.rm=TRUE)+
  geom_sf(data=atl_msa,fill='transparent',color='black',linewidth=0.5)+
  scale_fill_manual(values=c('lightblue','pink','lightgreen','lightyellow','blue','orange'),na.value = "transparent",name='Land Cover')+
  theme_void()+
  ggtitle('Land Cover in 2021')

)

```

Calculate the percent of each grid square that is developed in 2011 and 2021.

```{r land_cover_percent_developed, fig.width=10, fig.height=5}

lc2011_extract <- terra::extract(lc2011_class,fishnet,fun=table) %>%
  mutate(pct_developed = Developed / (Developed + Water + Forest + Farm + Wetlands + OtherUndeveloped) * 100)

lc2021_extract <- terra::extract(lc2021_class,fishnet,fun=table) %>%
  mutate(pct_developed2021 = Developed / (Developed + Water + Forest + Farm + Wetlands + OtherUndeveloped) * 100)

fishnet_pct_developed <- cbind(fishnet,lc2011_extract %>% select(pct_developed),lc2021_extract %>% select(pct_developed2021)) %>%
  mutate(change = pct_developed2021 - pct_developed)

grid.arrange(ncol=2,

ggplot(data=fishnet_pct_developed)+
  geom_sf(aes(fill=pct_developed),color='transparent')+
  scale_fill_viridis(option='rocket',name='Percent Developed')+
  geom_sf(data=atl_msa,fill='transparent',color='white',linewidth=0.5)+
  theme_void()+
  ggtitle('Percent Developed in 2011'),

ggplot(data=fishnet_pct_developed)+
  geom_sf(aes(fill=pct_developed2021),color='transparent')+
  scale_fill_viridis(option='rocket',name='Percent Developed')+
  geom_sf(data=atl_msa,fill='transparent',color='white',linewidth=0.5)+
  theme_void()+
  ggtitle('Percent Developed in 2021')
)
  
```

Determine grid squares that were developed in 2011 and 2021. For this analysis, consider any grid square that is more tha 25% developed to be developed.

``` {r identify_developed, fig.width=13, fig.height=7}

fishnet_pct_developed <- fishnet_pct_developed %>%
  mutate(developed2011 = as.factor(ifelse(pct_developed > 25,1,0)),
         developed2021 = as.factor(ifelse(pct_developed2021 > 25,1,0)),
         lc_change = as.factor(ifelse(developed2011 == 0 & developed2021 == 1 & change > 5,1,0)))

grid.arrange(ncol=3,

ggplot(data=fishnet_pct_developed)+
  geom_sf(aes(fill=developed2011),color='transparent')+
  scale_fill_manual(values=c('gray95','pink'),labels=c('Not Developed','Developed'),name='')+
  geom_sf(data=atl_msa,fill='transparent',color='black',linewidth=0.5)+
  theme_void()+
  ggtitle('Developed Pixels in 2011'),

ggplot(data=fishnet_pct_developed)+
  geom_sf(aes(fill=developed2021),color='transparent')+
  scale_fill_manual(values=c('gray95','pink'),labels=c('Not Developed','Developed'),name='')+
  geom_sf(data=atl_msa,fill='transparent',color='black',linewidth=0.5)+
  theme_void()+
  ggtitle('Developed Pixels in 2021'),

ggplot(data=fishnet_pct_developed)+
  geom_sf(aes(fill=lc_change),color='transparent')+
  scale_fill_manual(values=c('gray95','green'),labels=c('No Change','Change'),name='')+
  geom_sf(data=atl_msa,fill='transparent',color='black',linewidth=0.5)+
  theme_void()+
  ggtitle('Pixels with a Change in Development')
)

```

Determine the dominant land cover class in each grid square and also determine the percent of pixels in each grid square that are forest, farm, wetlands, and other. This was done so that we can test using both boolean and continuous variables for landcover in the model. 

```{r}

fishnet_2011 <- cbind(lc2011_extract,fishnet_pct_developed %>% select(developed2011,lc_change)) %>%
  rename(developed = developed2011) %>%
  rowwise() %>% mutate(max = max(c(Forest,Farm,Wetlands,Water,OtherUndeveloped,Developed))) %>%
  ungroup() %>%
  st_as_sf() %>%
  mutate(water=as.factor(ifelse(max==Water,1,0)),
         forest=as.factor(ifelse(max==Forest,1,0)),
         farm=as.factor(ifelse(max==Farm,1,0)),
         wetlands=as.factor(ifelse(max==Wetlands,1,0)),
         otherundeveloped=as.factor(ifelse(max==OtherUndeveloped,1,0)),
         pct_forest = Forest / (Developed + Water + Forest + Farm + Wetlands + OtherUndeveloped) * 100,
         pct_farm = Farm / (Developed + Water + Forest + Farm + Wetlands + OtherUndeveloped) * 100,
         pct_wetlands = Wetlands / (Developed + Water + Forest + Farm + Wetlands + OtherUndeveloped) * 100,
         pct_other = OtherUndeveloped / (Developed + Water + Forest + Farm + Wetlands + OtherUndeveloped) * 100) %>%
  select(ID,water,forest,farm,wetlands,otherundeveloped,lc_change,developed,pct_forest,pct_farm,pct_wetlands,pct_other,geometry)

```

## Spatial Lag of Development

I did spatial Lag to nearest 5 developed grid squares and 10 developed grid squares so that we can test models using both. 

``` {r}

fishnet_2011$lagDevelopment_5 <- nn_function(xyC(fishnet_2011),xyC(filter(fishnet_2011,developed==1)),5)
fishnet_2011$lagDevelopment_10 <- nn_function(xyC(fishnet_2011),xyC(filter(fishnet_2011,developed==1)),10)

ggplot()+
  geom_sf(data=fishnet_2011,aes(fill=lagDevelopment_10),color='transparent')+
  geom_sf(data=atl_msa,color='black',linewidth=0.5,fill='transparent')+
  scale_fill_viridis_c(option='virdis',name='Spatial Lag to Development (feet)')+
  labs(title = "Spatial Lag to 2011 Development")+
  theme_void()

```


## 2.4. Census Data

```{r load_key, warning = FALSE, eval = FALSE}
census_api_key("ab9309f9cc70c0e1895e7166c3ca981c40cf0331", overwrite = TRUE , install = TRUE)
```


```{r, warning = FALSE, message = FALSE, results = "hide"}
atlantaPop11 <- 
  get_acs(geography = "tract", variables = "B01003_001", year = 2011,
                state = 13, geometry = TRUE, 
                county=c("Fayette","Carroll","Rockdale","Cobb","Forsyth",
                         "Clayton","Henry","Dawson","Bartow","Lamar","Haralson","Meriwether","Newton","Gwinnett","Fulton","Pickens","Spalding","Douglas","Coweta","Heard","Butts","Jasper","DeKalb", "Cherokee","Walton","Pike","Barrow","Paulding")) %>%
  rename(pop_2011 = estimate) %>%
  st_transform(st_crs(atl_msa))

atlantaPop21 <- 
  get_acs(geography = "tract", variables = "B01003_001", year = 2021,
                state = 13, geometry = TRUE, 
                county=c("Fayette","Carroll","Rockdale","Cobb","Forsyth",
                         "Clayton","Henry","Dawson","Bartow","Lamar","Haralson","Meriwether","Newton","Gwinnett","Fulton","Pickens","Spalding","Douglas","Coweta","Heard","Butts","Jasper","DeKalb", "Cherokee","Walton","Pike","Barrow","Paulding")) %>%
  rename(pop_2021 = estimate) %>%
  st_transform(st_crs(atl_msa)) %>%
  st_buffer(-1)
```


# 2011 and 2021 census data plot

```{r, warning = FALSE, message = FALSE, fig.height= 8, fig.width= 11}
grid.arrange(
ggplot() +
  geom_sf(data = atlantaPop11, aes(fill=pop_2011), colour=NA) +
  scale_fill_viridis(option='rocket',name='Population 2011',direction=-1)+
  labs(title="Population, Atlanta MSA: 2011") +
  theme_void(),

ggplot() +
  geom_sf(data = atlantaPop21, aes(fill=pop_2021), colour=NA) +
  scale_fill_viridis(option='rocket',name='Population 2021',direction=-1)+
  labs(title="Population, Atlanta MSA: 2021") +
  theme_void(), ncol=2)
```


# Assigns a proportion of a tract’s population to a grid cell weighted by the proportion of the tract that intersects the grid cell

```{r, warning = FALSE, message = FALSE}

fishnet_2011 <-
  st_interpolate_aw(atlantaPop11["pop_2011"], fishnet_2011, extensive=TRUE) %>%
  st_drop_geometry() %>%
  rownames_to_column(var = "ID") %>%
  left_join(fishnet_2011 %>%
              mutate(ID = as.character(ID)),.,
            by=c('ID'='ID')) %>% 
  mutate(pop = replace_na(pop_2011,0))

```

## Calculate Distance from Road to Fishnet Centroid

``` {r}

highways <- st_read('./Data/Expressways_Atlanta_Region.geojson') %>% st_transform('EPSG:2240')

centroid <- fishnet_2011 %>%
  st_centroid()

nearest_feat <- st_nearest_feature(centroid,highways)

fishnet_2011$highway_dist <- as.double(st_distance(centroid, highways[nearest_feat,], by_element=TRUE))

ggplot()+
  geom_sf(data=fishnet_2011,aes(fill=highway_dist),color='transparent')+
  scale_fill_viridis_c(name='Distance to Highway feet')+
  geom_sf(data=highways,color='red')+
  theme_void()

```

## Join Counties

``` {r}

fishnet_2011 <- st_join(fishnet_2011,atl_msa %>% rename (county = NAME10) %>% select(county),largest=TRUE) %>%
  filter(water == 0) 
  
```
# Data Exploration

# Modelling

# Build Fishnet for 2021

## Identify Dominant Land Cover Class 

```{r}

fishnet_2021 <- cbind(lc2021_extract,fishnet_pct_developed %>% select(developed2021)) %>%
  rename(developed = developed2021) %>%
  rowwise() %>% mutate(max = max(c(Forest,Farm,Wetlands,Water,OtherUndeveloped))) %>%
  ungroup() %>%
  mutate(water=as.factor(ifelse(max==Water,1,0)),
         forest=as.factor(ifelse(max==Forest,1,0)),
         farm=as.factor(ifelse(max==Farm,1,0)),
         wetlands=as.factor(ifelse(max==Wetlands,1,0)),
         otherundeveloped=as.factor(ifelse(max==OtherUndeveloped,1,0)),
         pct_forest = Forest / (Developed + Water + Forest + Farm + Wetlands + OtherUndeveloped) * 100,
         pct_farm = Farm / (Developed + Water + Forest + Farm + Wetlands + OtherUndeveloped) * 100,
         pct_wetlands = Wetlands / (Developed + Water + Forest + Farm + Wetlands + OtherUndeveloped) * 100,
         pct_other = OtherUndeveloped / (Developed + Water + Forest + Farm + Wetlands + OtherUndeveloped) * 100) %>%
  select(ID,water,forest,farm,wetlands,otherundeveloped,developed,pct_forest,pct_farm,pct_wetlands,pct_other,geometry) %>%
  st_as_sf()

```


## Spatial Lag of Development and Road Distance

``` {r}

fishnet_2021$lagDevelopment_5 <- nn_function(xyC(fishnet_2021),xyC(filter(fishnet_2021,developed==1)),5)
fishnet_2021$lagDevelopment_10 <- nn_function(xyC(fishnet_2021),xyC(filter(fishnet_2021,developed==1)),10)

fishnet_2021$highway_dist <- as.double(st_distance(centroid, highways[nearest_feat,], by_element=TRUE))

```

## Add Population

```{r}

fishnet_2021 <-
  st_interpolate_aw(atlantaPop21["pop_2021"], fishnet_2021, extensive=TRUE) %>%
  st_drop_geometry() %>%
  rownames_to_column(var = "ID") %>%
  left_join(fishnet_2021 %>%
              mutate(ID = as.character(ID)),
            ., by=c("ID"='ID')) %>% 
  mutate(pop = replace_na(pop_2021,0))

```

## Join Counties

``` {r}

fishnet_2021 <- st_join(fishnet_2021,atl_msa %>% rename (county = NAME10) %>% select(county),largest=TRUE) %>%
  filter(water == 0) 
  
```
# Predict on 2021 Fishnet