---
title: "CPLN_Final_RB"
author: "Richard Barad"
date: "2024-04-24"
output: html_document
---

```{r setup, echo=FALSE}

library(sf)
library(tidycensus)
library(terra)
library(tidyverse)
library(tidyterra)
library(FNN)
library(gridExtra)

palette2 <- c("#41b6c4","#253494")
palette4 <- c("#a1dab4","#41b6c4","#2c7fb8","#253494")
palette5 <- c("#ffffcc","#a1dab4","#41b6c4","#2c7fb8","#253494")
palette10 <- c("#f7fcf0","#e0f3db","#ccebc5","#a8ddb5","#7bccc4",
               "#4eb3d3","#2b8cbe","#0868ac","#084081","#f7fcf0")

```

```{r functions, echo=FALSE}

xyC <- function(aPolygonSF) {
  as.data.frame(
    cbind(x=st_coordinates(st_centroid(aPolygonSF))[,1],
          y=st_coordinates(st_centroid(aPolygonSF))[,2]))
} 

nn_function <- function(measureFrom,measureTo,k) {
  #convert the sf layers to matrices
  measureFrom_Matrix <-
    as.matrix(measureFrom)
  measureTo_Matrix <-
    as.matrix(measureTo)
  nn <-   
    get.knnx(measureTo, measureFrom, k)$nn.dist
    output <-
    as.data.frame(nn) %>%
    rownames_to_column(var = "thisPoint") %>%
    gather(points, point_distance, V1:ncol(.)) %>%
    arrange(as.numeric(thisPoint)) %>%
    group_by(thisPoint) %>%
    summarize(pointDistance = mean(point_distance)) %>%
    arrange(as.numeric(thisPoint)) %>% 
    dplyr::select(-thisPoint) %>%
    pull()
  
  return(output)  
}

quintileBreaks <- function(df,variable) {
    as.character(quantile(df[[variable]],
                          c(.01,.2,.4,.6,.8),na.rm=T))
}
```

# Get Data

## Get MSA for Atlanta

```{r atl_msa}

atl_msa <- st_read('https://arcgis.atlantaregional.com/arcgis/rest/services/OpenData/FeatureServer/67/query?outFields=*&where=1%3D1&f=geojson') %>%
  st_make_valid() %>%
  st_transform('EPSG:2240') %>%
  dplyr::filter(MSA == 'Y')

atl_msa_dissolve <- st_union(atl_msa)

```

## Get Land Cover Data

```{r atl_landcover}

lc2011 <- rast('DATA/nlcd_2011.tif')
lc2011 <- project(lc2011,'EPSG:2240')
lc2011 <- terra::mask(lc2011,atl_msa)

lc2021 <- rast('DATA/nlcd_2021.tif')
lc2021 <- project(lc2021,'EPSG:2240')
lc2021 <- terra::mask(lc2021,atl_msa)

```

## Identify developed/undeveloped areas and change

```{r developed_area}

reclassMatrix <- 
  matrix(c(
    0,12,0,
    12,24,1,
    24,Inf,0),
  ncol=3, byrow=T)

developed_2011 <- classify(lc2011,reclassMatrix)

developed_2021 <- classify(lc2021,reclassMatrix)

development_change <- developed_2011+developed_2021

reclass2 <- matrix(c(0,0,
                     1,1,
                     2,0),ncol=2,byrow=T)

development_change <- classify(development_change,reclass2)

levels(development_change) <- c('No_Change','Change')

plot(development_change)

```
# Create Fishnet for 2011

```{r create_fishnet}

fishnet_2011 <- st_make_grid(atl_msa,4000) %>%
  st_sf() %>%
  st_join(.,atl_msa,predicate='intersects',left=FALSE) %>%
  mutate(ID=row_number()) %>%
  select(ID)
 
ggplot()+
  geom_sf(data=fishnet_2011,color='grey50',fill='gray90',linewidth=0.1)+
  geom_sf(data=atl_msa,fill='transparent',color='black',linewidth=0.5)+
  theme_void()

```

# Identify grid squares that are developed and not developed

```{r summarize_development}

dev_ch <- terra::extract(development_change,fishnet_2011,fun=table) %>%
  mutate(percent = Change / (No_Change + Change) * 100)

hist(dev_ch$percent,breaks=100)

fishnet_2011 <- dev_ch %>%
  mutate(lc_change = as.factor(ifelse(dev_ch$percent < 5,0,1))) %>%
  select(lc_change) %>%
  cbind(fishnet_2011,.) %>%
  drop_na(lc_change) %>%
  mutate(ID=row_number())

ggplot()+
  geom_sf(data=fishnet_2011,aes(fill=lc_change),color='transparent')+
  geom_sf(data=atl_msa,color='black',fill='transparent')+
  scale_fill_manual(values=c('gray95','green'),name='Development Change',labels=c('No Change','Change'))+
  theme_void()

```

# Identify Dominant Land Cover Class 

```{r, reclassify_dev, warning = FALSE, message = FALSE}

land_cover_reclass_matrix = matrix(
                           c(11,0,
                             21,1,
                             22,1,
                             23,1,
                             24,1,
                             41,2,
                             42,2,
                             43,2,
                             81,3,
                             82,3,
                             90,4,
                             95,4,
                             52,5,
                             71,5,
                             31,5),
                             ncol=2,byrow = TRUE)

lc2011_class <- terra::classify(lc2011,land_cover_reclass_matrix)

levels(lc2011_class) <- c('Water','Developed','Forest','Farm','Wetlands','OtherUndeveloped')

ggplot()+
  geom_spatraster(data=lc2011_class,na.rm=TRUE)+
  scale_fill_manual(values=c('lightblue','pink','lightgreen','lightyellow','blue','orange'),na.value = "transparent",name='Land Cover')+
  theme_void()

```

```{r}

lc2011_extract <- terra::extract(lc2011_class,fishnet_2011,fun=table) %>%
  pivot_longer(cols=-ID,names_to='Category',values_to='Count') 

onehop <- function(df,value) {
  df %>%
  mutate(!!value := ifelse(Category == {{value}},1,0))
}

fishnet_2011 <- lc2011_extract %>%
  group_by(ID) %>% summarize(Count = max(Count)) %>%
  left_join(.,lc2011_extract,by=c('ID','Count'),multiple='first') %>%
  select(ID,Category) %>%
  onehop(.,'Water') %>%
  onehop(.,'Developed') %>%
  onehop(.,'Forest') %>%
  onehop(.,'Farm') %>%
  onehop(.,'Wetlands') %>%
  onehop(.,'OtherUndeveloped') %>%
  mutate(ID = as.integer(ID)) %>%
  inner_join(.,fishnet_2011,by='ID') %>%
  select(-Category) %>%
  st_as_sf()

```

## Spatial Lag of Development

``` {r}

fishnet_2011$lagDevelopment_5 <- nn_function(xyC(fishnet_2011),xyC(filter(fishnet_2011,Developed==1)),5)
fishnet_2011$lagDevelopment_10 <- nn_function(xyC(fishnet_2011),xyC(filter(fishnet_2011,Developed==1)),10)

ggplot()+
  geom_sf(data=fishnet_2011,aes(fill=lagDevelopment_10),color='transparent')+
  geom_sf(data=atl_msa,color='black',linewidth=1,fill='transparent')+
  scale_fill_viridis_c(option='virdis',name='Spatial Lag to Development (feet)')+
  labs(title = "Spatial Lag to 2011 Development")+
  theme_void()

```

## Join Counties

``` {r}

fishnet_2011 <- st_join(fishnet_2011,atl_msa %>% rename (county = NAME10) %>% select(county),largest=TRUE) %>%
  filter(Water == 0) 
  
```
## Calculate Distance from Road to Fishnet Centroid

``` {r}

```

## 2.4. Census Data

```{r load_key, warning = FALSE, eval = FALSE}
census_api_key("ab9309f9cc70c0e1895e7166c3ca981c40cf0331", overwrite = TRUE , install = TRUE)
```


```{r, warning = FALSE, message = FALSE, results = "hide"}
atlantaPop11 <- 
  get_acs(geography = "tract", variables = "B01003_001", year = 2011,
                state = 13, geometry = TRUE, 
                county=c("Fayette","Carroll","Rockdale","Cobb","Forsyth",
                         "Clayton","Henry","Dawson","Bartow","Lamar","Haralson","Meriwether","Newton","Gwinnett","Fulton","Pickens","Spalding","Douglas","Coweta","Heard","Butts","Jasper","DeKalb", "Cherokee","Walton","Pike","Barrow","Paulding")) %>%
  rename(pop_2011 = estimate) %>%
  st_transform(st_crs(atl_msa))
```



```{r, warning = FALSE, message = FALSE, results = "hide"}
atlantaPop21 <- 
  get_acs(geography = "tract", variables = "B01003_001", year = 2021,
                state = 13, geometry = TRUE, 
                county=c("Fayette","Carroll","Rockdale","Cobb","Forsyth",
                         "Clayton","Henry","Dawson","Bartow","Lamar","Haralson","Meriwether","Newton","Gwinnett","Fulton","Pickens","Spalding","Douglas","Coweta","Heard","Butts","Jasper","DeKalb", "Cherokee","Walton","Pike","Barrow","Paulding")) %>%
  rename(pop_2021 = estimate) %>%
  st_transform(st_crs(atl_msa)) %>%
  st_buffer(-1)
```

# 2011 and 2021 census data plot

```{r, warning = FALSE, message = FALSE, fig.height= 8, fig.width= 11}
grid.arrange(
ggplot() +
  geom_sf(data = atlantaPop11, aes(fill=factor(ntile(pop_2011,5))), colour=NA) +
  scale_fill_manual(values = palette5,
                    labels=quintileBreaks(atlantaPop11,"pop_2011"),
                   name="Quintile\nBreaks") +
  labs(title="Population, Atlanta MSA: 2011") +
  theme_void(),

ggplot() +
  geom_sf(data = atlantaPop21, aes(fill=factor(ntile(pop_2021,5))), colour=NA) +
  scale_fill_manual(values = palette5,
                    labels=quintileBreaks(atlantaPop21,"pop_2021"),
                   name="Quintile\nBreaks") +
  labs(title="Population, Atlanta MSA: 2021") +
  theme_void(), ncol=2)
```


# Assigns a proportion of a tract’s population to a grid cell weighted by the proportion of the tract that intersects the grid cell

```{r, warning = FALSE, message = FALSE}

fishnet_2011 <-
  st_interpolate_aw(atlantaPop11["pop_2011"], fishnet_2011, extensive=TRUE) %>%
  st_drop_geometry() %>%
  rownames_to_column(var = "ID") %>%
  left_join(fishnet_2011 %>%
              mutate(ID = as.character(ID),.),
            by=c("ID"='ID')) %>% 
  mutate(pop_2011 = replace_na(pop_2011,0))

```


# Build Fishnet for 2021

```{r create_fishnet}

fishnet_2021 <- st_make_grid(atl_msa,4000) %>%
  st_sf() %>%
  st_join(.,atl_msa,predicate='intersects',left=FALSE) %>%
  mutate(ID=row_number()) %>%
  select(ID)

```

## Identify Dominant Land Cover Class 

```{r, reclassify_dev, warning = FALSE, message = FALSE}


lc2021_class <- terra::classify(lc2021,land_cover_reclass_matrix)

levels(lc2021_class) <- c('Water','Developed','Forest','Farm','Wetlands','OtherUndeveloped')

ggplot()+
  geom_spatraster(data=lc2021_class,na.rm=TRUE)+
  scale_fill_manual(values=c('lightblue','pink','lightgreen','lightyellow','blue','orange'),na.value = "transparent",name='Land Cover')+
  theme_void()

```

```{r}

lc2021_extract <- terra::extract(lc2021_class,fishnet_2021,fun=table) %>%
  pivot_longer(cols=-ID,names_to='Category',values_to='Count') 

onehop <- function(df,value) {
  df %>%
  mutate(!!value := ifelse(Category == {{value}},1,0))
}

fishnet_2021 <- lc2021_extract %>%
  group_by(ID) %>% summarize(Count = max(Count)) %>%
  left_join(.,lc2021_extract,by=c('ID','Count'),multiple='first') %>%
  select(ID,Category) %>%
  onehop(.,'Water') %>%
  onehop(.,'Developed') %>%
  onehop(.,'Forest') %>%
  onehop(.,'Farm') %>%
  onehop(.,'Wetlands') %>%
  onehop(.,'OtherUndeveloped') %>%
  mutate(ID = as.integer(ID)) %>%
  inner_join(.,fishnet_2021,by='ID') %>%
  select(-Category) %>%
  st_as_sf()
```


## Spatial Lag of Development

``` {r}

fishnet_2021$lagDevelopment_5 <- nn_function(xyC(fishnet_2021),xyC(filter(fishnet_2021,Developed==1)),5)
fishnet_2021$lagDevelopment_10 <- nn_function(xyC(fishnet_2021),xyC(filter(fishnet_2021,Developed==1)),10)

```

## Add Population

```{r}

fishnet_2021 <-
  st_interpolate_aw(atlantaPop21["pop_2021"], fishnet_2021, extensive=TRUE) %>%
  st_drop_geometry() %>%
  rownames_to_column(var = "ID") %>%
  left_join(fishnet_2021 %>%
              mutate(ID = as.character(ID)),
            ., by=c("ID"='ID')) %>% 
  mutate(pop_2021 = replace_na(pop_2021,0))

```

## Join Counties

``` {r}

fishnet_2021 <- st_join(fishnet_2021,atl_msa %>% rename (county = NAME10) %>% select(county),largest=TRUE) %>%
  filter(Water == 0) 
  
```
