---
title: "CPLN 675 - Urban Growth Modeling"
author: "Richard Barad"
date: "4/29/2024"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1.2. Setup

```{r load_packages, message=FALSE, warning=FALSE, results = "hide"}
library(tidyverse)
library(sf)
library(raster)
library(knitr)
library(kableExtra)
library(tidycensus)
library(tigris)
library(FNN)
#library(QuantPsyc) # JE Note: in R 4.1, QuantPsyc package not available.
library(caret)
library(yardstick)
library(pscl)
library(plotROC) 
library(ggrepel)
library(pROC)
library(grid)
library(gridExtra)
library(viridis)
library(igraph)
library(ggplot2)
library(terra)

plotTheme <- theme(
  plot.title =element_text(size=12),
  plot.subtitle = element_text(size=8),
  plot.caption = element_text(size = 6),
  axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 10),
  axis.title.y = element_text(size = 10),
  # Set the entire chart region to blank
  panel.background=element_blank(),
  plot.background=element_blank(),
  #panel.border=element_rect(colour="#F0F0F0"),
  # Format the grid
  panel.grid.major=element_line(colour="#D0D0D0",size=.75),
  axis.ticks=element_blank())

mapTheme <- theme(plot.title =element_text(size=12),
                  plot.subtitle = element_text(size=8),
                  plot.caption = element_text(size = 6),
                  axis.line=element_blank(),
                  axis.text.x=element_blank(),
                  axis.text.y=element_blank(),
                  axis.ticks=element_blank(),
                  axis.title.x=element_blank(),
                  axis.title.y=element_blank(),
                  panel.background=element_blank(),
                  panel.border=element_blank(),
                  panel.grid.major=element_line(colour = 'transparent'),
                  panel.grid.minor=element_blank(),
                  legend.direction = "vertical", 
                  legend.position = "right",
                  legend.key.height = unit(1, "cm"), legend.key.width = unit(0.2, "cm"))

palette2 <- c("#41b6c4","#253494")
palette4 <- c("#a1dab4","#41b6c4","#2c7fb8","#253494")
palette5 <- c("#ffffcc","#a1dab4","#41b6c4","#2c7fb8","#253494")
palette10 <- c("#f7fcf0","#e0f3db","#ccebc5","#a8ddb5","#7bccc4",
               "#4eb3d3","#2b8cbe","#0868ac","#084081","#f7fcf0")
```



```{r functions, echo=FALSE}

xyC <- function(aPolygonSF) {
  as.data.frame(
    cbind(x=st_coordinates(st_centroid(aPolygonSF))[,1],
          y=st_coordinates(st_centroid(aPolygonSF))[,2]))
} 

nn_function <- function(measureFrom,measureTo,k) {
  #convert the sf layers to matrices
  measureFrom_Matrix <-
    as.matrix(measureFrom)
  measureTo_Matrix <-
    as.matrix(measureTo)
  nn <-   
    get.knnx(measureTo, measureFrom, k)$nn.dist
    output <-
    as.data.frame(nn) %>%
    rownames_to_column(var = "thisPoint") %>%
    gather(points, point_distance, V1:ncol(.)) %>%
    arrange(as.numeric(thisPoint)) %>%
    group_by(thisPoint) %>%
    summarize(pointDistance = mean(point_distance)) %>%
    arrange(as.numeric(thisPoint)) %>% 
    dplyr::select(-thisPoint) %>%
    pull()
  
  return(output)  
}
```


# Helper Function


```{r, warning = FALSE, message = FALSE}
#this function converts a column in to quintiles. It is used for mapping.
quintileBreaks <- function(df,variable) {
    as.character(quantile(df[[variable]],
                          c(.01,.2,.4,.6,.8),na.rm=T))
}

#This function can be used to convert a polygon sf to centroids xy coords.
xyC <- function(aPolygonSF) {
  as.data.frame(
    cbind(x=st_coordinates(st_centroid(aPolygonSF))[,1],
          y=st_coordinates(st_centroid(aPolygonSF))[,2]))
} 

#this function convert a raster to a data frame so it can be plotted in ggplot
rast_2 <- function(inRaster) {
  data.frame(
    xyFromCell(inRaster, 1:ncell(inRaster)), 
    value = getValues(inRaster)) }
```

# 2. Data Wrangling & Feature Engineering


```{r load_data, warning = FALSE, message = FALSE, results = "hide"}
atlantaMSA <- 
  st_read("C:/Users/jonat/Documents/CPLN675/CPLN6750_Final/Counties_Atlanta_Region.geojson") %>%
  st_transform('EPSG:2240') %>%
  dplyr::filter(MSA == 'Y')

```


## 2.2. Land Cover Change Data


```{r atl_landcover}

lc2011 <- rast('DATA/nlcd_2011.tif')
lc2011 <- project(lc2011,'EPSG:2240')
lc2011 <- terra::mask(lc2011,atlantaMSA)

lc2021 <- rast('DATA/nlcd_2021.tif')
lc2021 <- project(lc2021,'EPSG:2240')
lc2021 <- terra::mask(lc2021,atlantaMSA)

```


## Identify developed/undeveloped areas and change

```{r developed_area}

reclassMatrix <- 
  matrix(c(
    0,12,0,
    12,24,1,
    24,Inf,0),
  ncol=3, byrow=T)

developed_2011 <- classify(lc2011,reclassMatrix)

developed_2021 <- classify(lc2021,reclassMatrix)

development_change <- developed_2011+developed_2021

reclass2 <- matrix(c(0,0,
                     1,1,
                     2,0),ncol=2,byrow=T)

development_change <- classify(development_change,reclass2)

levels(development_change) <- c('No_Change','Change')

plot(development_change)

```


```{r, warning = FALSE, message = FALSE}
atlantaMSA_fishnet <- 
  st_make_grid(atlantaMSA, 4000) %>%
  st_sf()

atlantaMSA_fishnet <-
  atlantaMSA_fishnet[atlantaMSA,]
```


```{r, warning = FALSE, message= FALSE}
ggplot() +
  geom_sf(data=atlantaMSA_fishnet) +
  labs(title="Fishnet, 4000 Foot Resolution") +
  mapTheme
```


## 2.3. Land Cover in 2001


## 2.4. Census Data

```{r load_key, warning = FALSE, eval = FALSE}
census_api_key("ab9309f9cc70c0e1895e7166c3ca981c40cf0331", overwrite = TRUE , install = TRUE)
```


```{r, warning = FALSE, message = FALSE, results = "hide"}
atlantaPop11 <- 
  get_acs(geography = "tract", variables = "B01003_001", year = 2011,
                state = 13, geometry = TRUE, 
                county=c("Fayette","Carroll","Rockdale","Cobb","Forsyth",
                         "Clayton","Henry","Dawson","Bartow","Lamar","Haralson","Meriwether","Newton","Gwinnett","Fulton","Pickens","Spalding","Douglas","Coweta","Heard","Butts","Jasper","DeKalb", "Cherokee","Walton","Pike","Barrow","Paulding")) %>%
  rename(pop_2011 = estimate) %>%
  st_transform(st_crs(atlantaMSA_fishnet))
```



```{r, warning = FALSE, message = FALSE, results = "hide"}
atlantaPop21 <- 
  get_acs(geography = "tract", variables = "B01003_001", year = 2021,
                state = 13, geometry = TRUE, 
                county=c("Fayette","Carroll","Rockdale","Cobb","Forsyth",
                         "Clayton","Henry","Dawson","Bartow","Lamar","Haralson","Meriwether","Newton","Gwinnett","Fulton","Pickens","Spalding","Douglas","Coweta","Heard","Butts","Jasper","DeKalb", "Cherokee","Walton","Pike","Barrow","Paulding")) %>%
  rename(pop_2021 = estimate) %>%
  st_transform(st_crs(atlantaMSA_fishnet)) %>%
  st_buffer(-1)
```

# 2011 and 2021 census data plot

```{r, warning = FALSE, message = FALSE, fig.height= 8, fig.width= 11}
grid.arrange(
ggplot() +
  geom_sf(data = atlantaPop11, aes(fill=factor(ntile(pop_2011,5))), colour=NA) +
  scale_fill_manual(values = palette5,
                    labels=quintileBreaks(atlantaPop11,"pop_2011"),
                   name="Quintile\nBreaks") +
  labs(title="Population, Atlanta MSA: 2011") +
  mapTheme,

ggplot() +
  geom_sf(data = atlantaPop21, aes(fill=factor(ntile(pop_2021,5))), colour=NA) +
  scale_fill_manual(values = palette5,
                    labels=quintileBreaks(atlantaPop21,"pop_2021"),
                   name="Quintile\nBreaks") +
  labs(title="Population, Atlanta MSA: 2021") +
  mapTheme, ncol=2)
```


# Assigns a proportion of a tract’s population to a grid cell weighted by the proportion of the tract that intersects the grid cell

```{r, warning = FALSE, message = FALSE}
atlantaMSA_fishnet <-
  atlantaMSA_fishnet %>%
  rownames_to_column("fishnetID") %>% 
  mutate(fishnetID = as.numeric(fishnetID)) %>%
  dplyr::select(fishnetID)

fishnetPopulation11 <-
  st_interpolate_aw(atlantaPop11["pop_2011"], atlantaMSA_fishnet, extensive=TRUE) %>%
  as.data.frame(.) %>%
  rownames_to_column(var = "fishnetID") %>%
  left_join(atlantaMSA_fishnet %>%
              mutate(fishnetID = as.character(fishnetID)),
            ., by=c("fishnetID"='fishnetID')) %>% 
  mutate(pop_2011 = replace_na(pop_2011,0)) %>%
  dplyr::select(pop_2011)

fishnetPopulation21 <-
  st_interpolate_aw(atlantaPop21["pop_2021"], atlantaMSA_fishnet, extensive=TRUE) %>%
  as.data.frame(.) %>%
  rownames_to_column(var = "fishnetID") %>%
  left_join(atlantaMSA_fishnet %>%
              mutate(fishnetID = as.character(fishnetID)),
            ., by=c("fishnetID"='fishnetID')) %>% 
  mutate(pop_2021 = replace_na(pop_2021,0)) %>%
  dplyr::select(pop_2021)

fishnetPopulation <- 
  cbind(fishnetPopulation11,fishnetPopulation21) %>%
  dplyr::select(pop_2011,pop_2021) %>%
  mutate(pop_Change = pop_2021 - pop_2011)
```

For comparison purposes, both the 2010 census tract geometries and the population weighted grid cells are plotted.

```{r, warning = FALSE, message = FALSE, fig.height = 8, fig.width= 11}
grid.arrange(
ggplot() +
  geom_sf(data=atlantaPop21, aes(fill=factor(ntile(pop_2021,5))),colour=NA) +
  scale_fill_manual(values = palette5,
                    labels=substr(quintileBreaks(atlantaPop21,"pop_2021"),1,4),
                   name="Quintile\nBreaks") +
  labs(title="Population, Atlanta MSA: 2021",
       subtitle="Represented as tracts; Boundaries omitted") +
  mapTheme,

ggplot() +
  geom_sf(data=fishnetPopulation, aes(fill=factor(ntile(pop_2021,5))),colour=NA) +
  scale_fill_manual(values = palette5,
                   labels=substr(quintileBreaks(fishnetPopulation,"pop_2021"),1,4),
                   name="Quintile\nBreaks") +
  labs(title="Population, Atlanta MSA: 2021",
       subtitle="Represented as fishnet gridcells; Boundaries omitted") +
  mapTheme, ncol=2)
```


## 2.5. Highway Distance


```{r, warning = FALSE, message = FALSE, results = "hide"}
atlantaHighways <-
  st_read("C:/Users/jonat/Documents/CPLN675/CPLN6750_Final/DATA/Expressways_Atlanta_Region.geojson") %>%
  st_transform(st_crs(atlantaMSA)) %>%
  st_intersection(atlantaMSA)
```

Below are some great r-based raster skills. The distance from each grid cell to its nearest highway segment is measured.

First, the highway layer is converted to raster. This is done by creating an `emptyRaster` of `NA` grid cells at the same spatial extent as `lc_change`. Then, `highway_raster` is created by converting `houstonHighways` to `sp` form and then to applying `rasterize`. The raster is then converted to points with `rasterToPoints` and `st_as_sf`, then `aggregate` is used to calculate mean distance by grid cell.

You may (but likely not) be interested in learning that `sp` is the older spatial data convention in R. Although `sf` is the new convention, raster/vector interactions still require `sp`. The `as` function converts.

```{r, warning = FALSE, message = FALSE}
development_change_raster <-raster(development_change)
emptyRaster <- development_change_raster
emptyRaster[] <- NA

highway_raster <- 
  as(atlantaHighways,'Spatial') %>%
  raster::rasterize(.,emptyRaster)

highway_raster_distance <- distance(highway_raster)
names(highway_raster_distance) <- "distance_highways"

highwayPoints <-
  rasterToPoints(highway_raster_distance) %>%
  as.data.frame() %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(atlantaMSA_fishnet))

highwayPoints_fishnet <- 
  aggregate(highwayPoints, atlantaMSA_fishnet, mean) %>%
  mutate(distance_highways = ifelse(is.na(distance_highways),0,distance_highways))

ggplot() +
  geom_sf(data=atlantaMSA) +
  geom_point(data=highwayPoints_fishnet, aes(x=xyC(highwayPoints_fishnet)[,1], 
                                             y=xyC(highwayPoints_fishnet)[,2], 
                 colour=factor(ntile(distance_highways,5))),size=1.5) +
  scale_colour_manual(values = palette5,
                      labels=substr(quintileBreaks(highwayPoints_fishnet,"distance_highways"),1,8),
                      name="Quintile\nBreaks") +
  geom_sf(data=atlantaHighways, colour = "red") +
  labs(title = "Distance to Highways",
       subtitle = "As fishnet centroids; Highways visualized in red") +
  mapTheme
```

1. plot of each land cover class. how many 0 and 1


